Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-08 numerical.tex
edits to 6.4.  tried to more clearly state goal at outset and remove some repetitive discussion that repeats 6.3.  added speculation at end for non-polytropes.  I guess I'm officially happy now.

%we plot the maximum VSI growth rates
%found in the fiducial disk model as a function of the thermal
%relaxation timescale $\beta$ for several values of $\khat$ and
%$\zmax$. We also plot the critical timescale
%$\beta_\mathrm{crit}=0.125$ for the fiducial disk parameters.   
%
%Note that the curves are not neccessarily smooth because the most unstable
%mode may not always correspond to the same type of disturbance. For example,
%as seen above, for $\khat \leq 10$ higher order body modes are
%dominant as $\beta\to 0$, but the fundamental 
%mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
%can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
%$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
%$\khat=5,\,10$ (and $30$ for $\zmax=3H$), we find
%$\beta_\mathrm{crit}$ accurately predicts the thermal timescal beyond
%which the VSI is suppressed.   
%
%Complications from vertical boundaries arise when considering large
%$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
%0.1$ the dominant disturbance correspond to the leftmost eigenvalues 
%seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
%which do not exist for $\zmax=3H$. For $\khat=50,\,100$ and
%$\beta\to0$, we find surface modes dominate so the maximum growth rate
%depends on $\zmax$ in the limit of rapid thermal relaxation. 
%However, the maximum growth rate is insensitive to 
%$\zmax$ for moderately large $\beta$ ($\gtrsim 0.1$), i.e. close to
%the stabilization of the fundamental mode.

%Our result is consistent with numerical simulations performed by
%\citetalias{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
%0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
%rates for $\beta\gtrsim 0.6$ are small, and are restricted to very
%small radial lengthscales $\sim 10^{-2}H$, for which current numerical
%simulations may not be able to resolve.
%
%As pointed out by \citetalias{barker15}, in vertically isothermal disks with
%an imposed boundary, surface modes are dependent on $\zmax$, but they
%are often the most unstable. Despite this caveat, and that surface
%modes are not accounted in our analysis, we see from
%Fig. \ref{bcrit_compare1} that $\beta_\mathrm{crit}$ nevertheless
%provides a reasonable estimate for an upper limit to the thermal
%timescale for the VSI.  
%
%We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
%(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
%point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
%$ h\in[0.02,0.1]$ separately. The numerically obtained
%$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
%comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
%result, together with Fig. \ref{bcrit_compare1}, shows that  
%$\beta \lesssim \beta_\mathrm{crit}$ is a practical
%criteria for the VSI to operate effectively in stably stratified,
%vertically isothermal disks.

\subsection{Vertically non-isothermal disk model} 
We briefly examine a vertically non-isothermal disk model with 
a physical surface. Recall from Eq. \ref{vertical_shear_ex} that in
general the vertical shear rate $\p_z\Omega\propto s$, the radial
entropy gradient. From the expression for the critical cooling timescale,
Eq. \ref{iso_vsi_cond}, it is reasonable to expect a similar criteria
for vertically non-isothermal disks in the form 
\begin{align}\label{bcrit_noniso}
 \beta_\mathrm{crit}\to\beta_\mathrm{crit, noniso} =
 \frac{h|s|}{\gamma - \Gamma}. 
\end{align}

Here, we consider a disk model with $(p,q, h)=(-0.5,-1,0.05)$ and $(\gamma,
\Gamma)=(1.4,1.3)$. We set the vertical domain size close to the
physical disk surface, $\zmax = 0.99H_s\simeq 2.6H$, which is smaller
than our standard domain size. For this setup, $s=-0.85$ and
$\beta_\mathrm{crit,noniso}=0.425$, which is $\sim 3$ times larger than
our fiducial, vertically isothermal disk with $\beta_\mathrm{crit}=0.125$.

% A disk surface at finite height
% means that higher order body modes and surface modes are better
% defined (cf. in vertically isothermal disk models where the disk
% boundary $\zmax$ is a free parameter, see \citetalias{barker15}).

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=30$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely suppressed. (Note the significant reduction in
growth rates for $\beta=0.4$, which is close to
$\beta_\mathrm{crit,noniso}$.) However, this behavior is not seen for 
large wavenumbers (see below).

@@

In Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model, which is
qualitatively similar to Fig. \ref{bcrit_compare1} for the vertically
isothermal disk. For $\khat=O(10)$, there is a   
precise cut-off thermal timescale close to that given by
Eq. \ref{bcrit_noniso}. We found that very large wavenumbers, 
$\khat > O(10^2)$, were needed to enable surface modes, and modes with
such values of $\khat$ do not have a maximum cooling time. However,
these small lengthscales may not be important in practice
(see \S\ref{caveats_visc}). Thus, $\beta<\beta_\mathrm{crit,noniso}$ 
provides an excellent criteria for instability.

@@

The above calculation demonstrates that the effect of introducing
finite cooling on the VSI operates very similarly in vertically
isothermal disks with an imposed surface, and in vertically
non-isothermal disks with a physical surface. This is despite the fact
that the vertical buoyancy frequency, $N_z$, formally diverges as $|z|\to H_s$.

>>>>

\subsection{Vertically non-isothermal disk model} 
We now generalize our critical cooling time by considering disks which
are not vertically isothermal.  We rely on physical arguments for this generalization.
In disks with weaker vertical stratification, i.e.\ $\Gamma > 1$, we expect $\beta_\mathrm{crit}$ to be larger.
We thus transform $1/(\gamma -1) \to 1/(\gamma - \Gamma)$.

We also expect $\beta_\mathrm{crit}$ to scale with the vertical shear.   
In general the vertical shear rate $\p_z\Omega\propto s$, the radial
entropy gradient, see Eq.\ \ref{vertical_shear_ex}.  We thus 
 transform $q \to s = q +p(1-\Gamma)$.

Our estimate for the generalized cooling 
 time criterion -- valid for both vertically isothermal and non-isothermal disks (cf.\ Eq.\ \ref{iso_vsi_cond}) -- is thus
\begin{align}\label{bcrit_noniso}
 \beta_\mathrm{crit}\to\beta_\mathrm{crit, noniso} =
 \frac{h|s|}{\gamma - \Gamma}. 
\end{align}

We test $\beta_\mathrm{crit, noniso}$ using a disk model with 
$(p,q, h)=(-0.5,-1,0.05)$ and $(\gamma,
\Gamma)=(1.4,1.3)$. For this setup, $s=-0.85$ and
$\beta_\mathrm{crit,noniso}=0.425$, which is $\sim 3$ times larger than
our fiducial, vertically isothermal disk with $\beta_\mathrm{crit}=0.125$.

We set the vertical domain size just inside the
physical, zero-density disk surface, $\zmax = 0.99H_s\simeq 2.6H$, see Eq.\ \ref{eqm_dens}. 
At this surface, the vertical buoyancy frequency, $N_z$,  diverges as $|z|\to H_s$.  While we might expect
pathological behavior for this model, we fortunately do not find it.

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=30$ and several values of $\beta$. This plot is 
qualitatively similar to the vertically isothermal case, Fig. \ref{compare_modes_cool_kx10}. As before, 
larger $\beta$ values rapidly
stabilize higher order modes.  For sufficiently large $\beta$ only the fundamental mode is unstable, with a growth rate
that decreases as $\beta$ approaches 
$\beta_\mathrm{crit,noniso}$.

@@

Fig.\ \ref{bcrit_compare2} shows how maximum VSI growth
rates vary with $\beta$ in the vertically non-isothermal model. The qualitative
behavior similar to Fig.\ \ref{bcrit_compare1} for the vertically
isothermal disk. For $\khat=O(10)$, growth rates drop to zero at the
predicted $\beta_\mathrm{crit,noniso}$.  As in the vertically isothermal case,
modes with both lower and higher wavenumbers  exhibit slow growth rates 
beyond the critical cooling time. As argued in \S\ref{bcrit_num_test},
the critical cooling time is useful despite not being an absolute stability limit.

@@

The generalized cooling time criterion in Eq.\ \ref{bcrit_noniso}
thus appears to be valid.  If a disk's vertical structure is
not characterized by a single polytropic index, $\Gamma$, we 
expect that a density weighted average of  $\beta_\mathrm{crit,noniso}$
should be a good approximation.  This expectation remains to be tested.
